Causal graph analysis of COVID-19 observational data in German districts reveals effects of determining factors on reported case numbers

### 2021-03-21 manuscript/v4_medrxiv_revision/discussion.Rmd
add reference causal collinearity, and discussion points

We observed overdispersion and a substantial increase in model performance with a negative binomial regression compared to Poisson regression, which is in line with the results on COVID-19 daily case counts of @kraemer_effect_2020 and others [@liu_impact_2020;@bannister-tyrrell_preliminary_2020;@Islamm2743]. We did not model case counts with a differential equation model like the classic SIR-model [@kermack1991contributions] and its successors, since these are more suited to prediction [e.g. @rkimodellierung] while our choice of a negative binomial regression framework allowed us to estimate the effects of confounders more reliably. There are more advanced statistical methods for count data, e.g. zero-inflated models and mixed models. We tested both approaches as extensions to the negative binomial regression and experienced numerical problems and increased computing time, along with an insubstantial increase in model performance. Furthermore, our model assumed that all variables have effects proportional to the size of their measurements. It is possible that some variables show saturation effects or opposite effects for low, medium, or high values. This could be modeled with polynomial or other transformations of the variables, which we did not employ due to limited temporal and spatial data availability. Use of a fixed DAG with effect estimation via regression assumes that data was generated by the same underlying process for the observation period. By inclusion of the successive mitigation interventions as binary variables we were able to explain some of the variance caused by the changing dynamics of case numbers (similar to [@Islamm2743]).

>>>>

We observed overdispersion and a substantial increase in model performance with a negative binomial regression compared to Poisson regression, which is in line with the results on COVID-19 daily case counts of @kraemer_effect_2020 and others [@liu_impact_2020;@bannister-tyrrell_preliminary_2020;@Islamm2743]. We did not model case counts with a differential equation model like the classic SIR-model [@kermack1991contributions] and its successors, since these are more suited to prediction [e.g. @rkimodellierung] while our choice of a negative binomial regression framework allowed us to estimate the effects of confounders more reliably. There are more advanced statistical methods for count data, e.g. zero-inflated models and mixed models. We tested both approaches as extensions to the negative binomial regression and experienced numerical problems and increased computing time, along with an insubstantial increase in model performance. Furthermore, our model assumed that all variables have effects proportional to the size of their measurements. It is possible that some variables show saturation effects or opposite effects for low, medium, or high values. This could be modeled with polynomial or other transformations of the variables, which we did not employ due to limited temporal and spatial data availability. Interaction effects of variables and confounding effects or mediating variables are explicitly taken care of by deriving the valid adjustment sets for a given exposure based on the causal DAG. Use of a fixed DAG with effect estimation via regression assumes that data was generated by the same underlying process for the observation period. By inclusion of the successive mitigation interventions as binary variables we were able to explain some of the variance caused by the changing dynamics of case numbers (similar to [@Islamm2743]). While multicollinearity of variables poses less of a problem for a proper causal graph analysis [@schisterman2017collinearitycausal], we addressed the problem of multicollinearity in our predictors by two approaches: principal component analysis for the highly collinear mobility variables as well as a regularized regression approach (ridge regression). The latter (in conjunction with cross-validation) also reduced the problem of overfitting.
